Interpretable Neural Networks using EAGGA

### 2025-03-15 paper/main.tex
appendix after references
own extension architectural details

  * group structures + datasets
    - group structures implemented as described in the paper with additional list encoding sign of monotonicity of the individual features as detected by the
      monotonicity detector
    - included features are passed to dataset, dataset outputs included features, multiplied by the features' individual monotonicity (-1 / 1), so that group monotonicity
      can be encoded with only $\{0,1\}$, as proposed in paper
    - entire group structure is passed to NN, where architecture is built accordingly

@@

  * neural network
    - "instead of XGB as in the original paper, we apply the method to NNs to examine whether this type of regularisation can achieve interpretability on NNs while
      outperforming EAGGA on XGB (original paper), ..., this requires special architecture, etc. pp."
    - NN
      * comprises "sub-NNs", one for each equivalence relation -> basically non-fully connected NN
      * hidden layers use ReLU activation and dropout afterwards, then a shared output layer with concatenated sub-NNs' outputs as input and sigmoid activation
      * optimizer is AdamW with default params, loss is binary cross entropy loss (sigmoid thus implicit in the loss function for better numerical stability,
        NN itself outputs "logits", which refers to pre-activation output in pytorch)
    - feature sparsity thus achieved by only training on included feature groups -> goes somewhat against notion of deep learning where model is supposed to decide itself,
      which feature to "use" / put importance on -> ELABORATE
    - feature interaction achieved by grouping, max-operation in ReLU induces interaction effect (different kind of interaction than e.g. in LM with multiplication)
      -> equation why the interacting features need to be grouped together when using ReLU, cf. photos
      * given layer input $\boldsymbol{x}$ and w.l.o.g. omitting bias
      * $\text{ReLU}(\boldsymbol{x})=\max(0,\sum_{j=1}^p w_p x_p)$
      * interaction not of the form $x_j \cdot x_{j'}$ as used to from LMs but of a form where the sum $w_j x_j+w_{j'} x_{j'}$ decides whether $x_j,x_{j'}$
        pass through the activation or not
      * in other words: $\max(0,w_j x_j + w_{j'} x_{j'})\neq\max(0,w_j x_j)+\max(w_{j'} x_{j'})$ and thus there is an interaction
    - monotonicty constraint achieved by clipping weights to [0, infty) after each epoch for restricted equivalence relations (monotonic decrease achieved
      via dataset object multiplying features with their individual signs), bias clipping not necessary (constant additive term)
      -> equation how monotonicity is enforced with this
      * consider vector form computation of hidden neuron $\boldsymbol{o}_{\text{in}}^{(l)}\in\mathbb{R}^{p^{(l)}}$ of layer $l$
        - $\boldsymbol{o}_{\text{in}}^{(l)}=\boldsymbol{W} \boldsymbol{o}_{\text{out}}^{(l-1)} + \boldsymbol{b}$
        - where $\boldsymbol{o}_{\text{in}}^{(l)}$ refers to pre-activation values, $\boldsymbol{o}_{\text{out}}^{(l)}$ to value after applying ReLU
        - weight matrix $\boldsymbol{W}\in\mathbb{R}^{p^{(l)} \times p^{(l-1)}}$
        - bias $\boldsymbol{b}\in\mathbb{R}^{p^{(l)}}$
      * definition of monotonically increasing function $f$: for $x_1\le x_2$ it must hold that $f(x_1)\le f(x_2)$
      * $\boldsymbol{W} \boldsymbol{o}_{\text{out}}^{(l-1)} + \boldsymbol{b}$ clearly only monotonic if and only if $\boldsymbol{W}\in\mathbb{R}_{0+}{p^{(l-1)}\times p^{(l)}}$
        - where $\mathbb{R}_{0+}=[0,\infty)$
        - hence weight clipping necessary, bias clipping not because only constant additive term
      * $\boldsymbol{o}_{\text{out}}^{(l)}=ReLU(\boldsymbol{o}_{\text{in}}^{(l)})=\max(0,\boldsymbol{o}_{\text{in}}^{(l)})$ monotonic
  (experimental setup)
  \includegraphics[width=\linewidth]{./dataset_split.png}
  \label{fig-dataset-split}
  * evaluation, holdout, cv, early stopping
    - evaluation via dominated hypervolume along AUC-ROC, NF, NI, NNM as defined in original paper
      * NF simply rel. \# of included features
      * NI = sum of all possible pairwise interactions in each group over all possible pairwise interactions among all features = $\frac{\sum_g^G {p_g \choose 2}}{{p \choose 2}}$
      * NNM = rel. \# unconstrained features
    - outer holdout split: 2/3 train, 1/3 test (as in paper)
      * run EAGGA on holdout train split, i.e. train + select best $\mu=100$ individuals based on non-dom-sorting (ascending) + crowding distance (descending) as tie breaker
    - inner CV split on outer train portion: 5-fold (as in paper)
      * in each fold fit model on 80\% of CV-train portion, use remaining 20\% of CV-train for early stopping
    - early stopping criterion
      * for each fold, always train for min 200 epochs, keep track of model with lowest loss
      * after that, use patience of 100: if current model's loss is > mean of last 100 epochs' losses -> stop early
      * if no early stopping, train each fold for max 10 minutes
      * then stop training, return model with lowest loss
    -> refere Figure \ref{fig-dataset-split}
    - then evaluate on last remaining from CV + keep best $\mu$ individuals + generate $\lambda=10$ offspring for new generation + repeat
    - final evaluation of individuals of pareto set
      * for each individual, train model on training set of hold-out for max \# of epochs taken during CV-training
      * decided on max \# epochs instead of mean after looking at loss graphs in preliminary experiments, also refer Figure \ref{fig-es-losses}
      * those show that there is no uptick in losses on the early stopping portion (disjunct from training portion) of the set, hence max is reasonable
  * hardware: training on Sagemaker Notebook instance
    - initial experiments on ml.g4dn.xlarge instance (2 vCPUs, 16GiB RAM, 1 NVIDIA T4, cf. https://aws.amazon.com/de/ec2/instance-types/g4/)
      using cuda not much faster than on ml.t3.medium (2 Intel Xeon 8000 vCPUs, 4GiB RAM, cf. https://aws.amazon.com/de/ec2/instance-types/t3/)
    - thus decided for more economic + ressourcen-schonend t3.medium
  * did not train on philippine and gina datasets (308, 970 features, respectively), as they ran out of memory when computing interaction detectors
  * known bugs
    - in rare cases (anecdotally once every ~5-10 datasets), gga\_mutate seems to be generating -1 as monotonicity attribute, despite np.random.randint(low=0, high=2, size=1)
      * loop crashes at group\_structure creation, for these cases subtract previous runtime from 8hrs + load from last generation in output via load\_population
      * so far only happened for madeline in after gen-8.json after 5hrs 45mins (at start of gen-9, which wasn't exported yet) -> loaded this and ran for another 2hrs 15mins
  * only 18 datasets

>>>>

We implement the group structures as described by \citet[p. 541]{EAGGA} with an additional list encoding whether the signs of individual features should
be inverted (-1) or not (1), as detected by the monotonicity detector.
The features included in the group structure are then passed to a custom pytorch Dataset implementation, which outputs only the included features, multiplied with their
individual sign. This allows the groups' monotonicity constraints to be encoded with only $\{0,1\}$, as proposed by \citet[p. 543]{EAGGA}.
The entire group structure is then passed to the neural network, where the architecture is built accordingly.

@@

Applying EAGGA to NNs requires defining a special architecture to realise interaction and monotonicity constraints.
The neural network consists of a group of fully-connected sub-networks that are not connected amongst each other, as visualised in Figure \ref{fig-nn-architecture}.
Each equivalence class in the group structure has its own sub-network, so that it is guaranteed that only these features can interact with each other.
For the hidden layers, we use ReLU activations with dropout afterwards, after which a shared output layer transforms the concatenated output of the sub-networks
to a probability using the Sigmoid activation function.
We use AdamW with default parameters, optimizing a binary cross-entropy loss with Sigmoid implicit in the loss function for better numerical stability. The network
itself thus outputs only a ``logit'', which is the pre-activation output in pytorch.
Feature sparsity is achieved by only training on the included features that are passed to the network from the dataset.
Grouping the network into sub-networks enables restricting feature interactions to only those features included in the respective group.
The $\max$-operation in the ReLU induces a certain interaction effect on its inputs. It is important to note that this is a different kind of interaction
than the multiplicative one happening for instance in a linear model.
Given a layer's input $\boldsymbol{x}$ and without loss of generality omitting the bias, the ReLU activation is defined as follows
  \text{ReLU}(\boldsymbol{x})=\max(0,\sum_{j=1}^p w_j x_j)
  \label{eq-our-extension-interaction-1}
We see that the interaction is not of the form $x_j \cdot x_{j'}$ but rather of a form where the sum $w_j x_j+w_{j'} x_{j'}$ decides whether $x_j,x_{j'}$
pass through the activation or not.
In other words, the interaction is given by the fact that
  \max(0,w_j x_j + w_{j'} x_{j'})\neq\max(0,w_j x_j)+\max(w_{j'} x_{j'})
  \label{eq-our-extension-interaction-2}
Finally, monotonicty is achieved by clipping the respective sub-network's weights to $[0,\infty)$ after each epoch. Whether a feature's effect is
increasing or decreasing depends on the individual feature's sign encoded in the custom dataset implementation.
Consider the vector form computation of a pre-activation hidden neuron $\boldsymbol{o}_{\text{in}}^{(l)}\in\mathbb{R}^{p^{(l)}}$ of layer $l$
  \boldsymbol{o}_{\text{in}}^{(l)}=\boldsymbol{W} \boldsymbol{o}_{\text{out}}^{(l-1)} + \boldsymbol{b}
  \label{eq-our-extension-monotonicity}
where $\boldsymbol{o}_{\text{out}}^{(l)}$ denotes the value after applying ReLU, $\boldsymbol{W}\in\mathbb{R}^{p^{(l)} \times p^{(l-1)}}$ refers to
the weight matrix and $\boldsymbol{b}\in\mathbb{R}^{p^{(l)}}$ to the bias.
A function $f$ is monotonically increasing if and only if $\forall x_1\le x_2$ it holds that $f(x_1)\le f(x_2)$ and thus
$\boldsymbol{W} \boldsymbol{o}_{\text{out}}^{(l-1)} + \boldsymbol{b}$ is clearly only monotonically increasing if and only if
$\boldsymbol{W}\in\mathbb{R}_{0+}^{p^{(l-1)}\times p^{(l)}}$ where $\mathbb{R}_{0+}=[0,\infty)$.
$\boldsymbol{o}_{\text{out}}^{(l)}=ReLU(\boldsymbol{o}_{\text{in}}^{(l)})=\max(0,\boldsymbol{o}_{\text{in}}^{(l)})$ Is therefore a monotonically increasing
function if and only if the weights are clipped as described. Bias clipping is not necessary as it is a constant additive term.

\subsection{Experimental Setup}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{./dataset_split.png}
  \caption{TODO: caption}
  \label{fig-dataset-split}
\end{figure}
  * evaluation, holdout, cv, early stopping
    - evaluation via dominated hypervolume along AUC-ROC, NF, NI, NNM as defined in original paper
      * NF simply rel. \# of included features
      * NI = sum of all possible pairwise interactions in each group over all possible pairwise interactions among all features = $\frac{\sum_g^G {p_g \choose 2}}{{p \choose 2}}$
      * NNM = rel. \# unconstrained features
    - outer holdout split: 2/3 train, 1/3 test (as in paper)
      * run EAGGA on holdout train split, i.e. train + select best $\mu=100$ individuals based on non-dom-sorting (ascending) + crowding distance (descending) as tie breaker
    - inner CV split on outer train portion: 5-fold (as in paper)
      * in each fold fit model on 80\% of CV-train portion, use remaining 20\% of CV-train for early stopping
    - early stopping criterion
      * for each fold, always train for min 200 epochs, keep track of model with lowest loss
      * after that, use patience of 100: if current model's loss is > mean of last 100 epochs' losses -> stop early
      * if no early stopping, train each fold for max 10 minutes
      * then stop training, return model with lowest loss
    -> refere Figure \ref{fig-dataset-split}
    - then evaluate on last remaining from CV + keep best $\mu$ individuals + generate $\lambda=10$ offspring for new generation + repeat
    - final evaluation of individuals of pareto set
      * for each individual, train model on training set of hold-out for max \# of epochs taken during CV-training
      * decided on max \# epochs instead of mean after looking at loss graphs in preliminary experiments, also refer Figure \ref{fig-es-losses}
      * those show that there is no uptick in losses on the early stopping portion (disjunct from training portion) of the set, hence max is reasonable
  * hardware: training on Sagemaker Notebook instance
    - initial experiments on ml.g4dn.xlarge instance (2 vCPUs, 16GiB RAM, 1 NVIDIA T4, cf. https://aws.amazon.com/de/ec2/instance-types/g4/)
      using cuda not much faster than on ml.t3.medium (2 Intel Xeon 8000 vCPUs, 4GiB RAM, cf. https://aws.amazon.com/de/ec2/instance-types/t3/)
    - thus decided for more economic + ressourcen-schonend t3.medium
  * did not train on philippine and gina datasets (308, 970 features, respectively), as they ran out of memory when computing interaction detectors
  * known bugs
    - in rare cases (anecdotally once every ~5-10 datasets), gga\_mutate seems to be generating -1 as monotonicity attribute, despite np.random.randint(low=0, high=2, size=1)
      * loop crashes at group\_structure creation, for these cases subtract previous runtime from 8hrs + load from last generation in output via load\_population
      * so far only happened for madeline in after gen-8.json after 5hrs 45mins (at start of gen-9, which wasn't exported yet) -> loaded this and ran for another 2hrs 15mins
  * only 18 datasets
